SFJ: An implementation of Semantic Featherweight Java

### 2020-02-28 conference/paper.tex
links and final pass

Consider the set of polygons, such as triangles, squares and rhombuses given by a class hierarchy.
We want to define a method \emph{diagonal} that takes a polygon and returns the length of its longest diagonal. This method makes sense only if the polygon passed to it has at least four sides, hence triangles are excluded.
In Java this could be implemented in the following ways:
$$
\begin{array}{l}
\synclass Polygon \ \{ \ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Other\_Polygons}{Polygon} {
\\
	\qquad \mathbf{real}\ diagonal(Other\_Polygons \ p)\ \{\ldots \}
	\\
	\hspace{3.5cm} \ldots
	\\
}
\\
\syndecl{Square}{Other\_Polygons}{\ldots}
\\
\syndecl{Rhombus}{Other\_Polygons} {\ldots}
\end{array}
$$
Or by means of an interface \emph{Diagonal}:
$$
\begin{array}{l}
public \ \syninterface Diagonal\ \{
	\\
	\qquad \mathbf{real}\ diagonal(Polygon \ p);
	\\
\}
\\
\synclass Polygon \ \{\ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\synidecl{Square}{Polygon} \synimpl Diagonal \ \{\ldots\}
\\
\synidecl{Rhombus}{Polygon} \synimpl Diagonal \ \{\ldots\}
\\
\hspace{4.5cm} \ldots
\end{array}
$$

@@

We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:
$$
\begin{array}{l}
\syndecl{Polygon}{Object} {\ldots}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Square}{Polygon} {\ldots}
\\
\syndecl{Rhombus}{Polygon} {\ldots}
\\
\hspace{2.5cm} \ldots
\\
\syndecl{Diagonal}{Object} {
	\\
	\qquad \mathbf{real}\ diagonal(\textit{Polygon} \textbf{ and not }  \textit{Triangle} \ p)\{\ldots\}
	\\
}
\end{array}
$$
We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

@@

if all classes are not resolved:
    program contains invalid type definition
\end{verbatim}
%If the types in the program are finite trees whose leaves are constants with no cycles, then at each iteration of the algorithm, either all the types are going to be resolved, or we are going to be able to resolve at least one type. 
%If we do not resolve at least one type and not all types are resolved, we know we have encountered a cycle in the type definition. 
%
Now, we can define the subtyping relation.
Building upon the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the subtyping relation by defining a map from a type to the set of its subtypes, with the property that the set of values of a subtype is included in the set of values of the type.
As a first step, we start with basic types.
\setlength{\abovedisplayskip}{5pt}
\setlength{\belowdisplayskip}{5pt}
\begin{flalign*}
    &Double  = \left\{Double, Float, Int, Short, Byte\right\} &&Float = \left\{Float, Short, Byte\right\}\\
    &Long = \left\{Long, Int, Short, Byte\right\}  &&Int = \left\{Int, Short, Byte\right\}\\
    &Short = \left\{Short, Byte\right\} && Byte = \left\{Byte\right\}\\
    &Boolean = \left\{Boolean \right\} 
\end{flalign*}
%
Note that \emph{Int} is not a subtype of \emph{Float} as a 32-bit \emph{float} cannot represent the whole set of 32-bit \emph{integer} values accurately and therefore \emph{Int} is not fully set-contained, however this is not the case for \emph{Int} and \emph{Double}. 
Similarly, \emph{Long} is not a subtype of \emph{Double}.

@@

function checkSuperSet(class, other):
    boolean flag = true
    for field in class:
        if field contains type not in relation:
            return false
        if other does not have field:
            flag = false
        else:
            if other.field.types not fully contains field.types:
                flag = false
    for method in class:
        if method contains type not in relation:
            return false
        if other does not have method:
            flag = false
        else:
            if other.method.types not fully contains method.types:
                flag = false
    if flag == true:
        add class to other subtype relation
\end{verbatim}
Let us illustrate the subtyping algorithm with our \emph{Polygons} example \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.
\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} && Triangle  = \left\{Triangle\right\}\\ 
    &Square = \left\{Square\right\} &&
    Diagonal = \left\{Diagonal\right\} 
\end{flalign*}
Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons: 
\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\} 
\end{flalign*}
If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.

The subtyping algorithm finds all nominal and structural subtypes.
It will find all nominal subtypes, as nominal subtypes inherit all fields and methods of their super class, so are always guaranteed to be a superset. 
Similarly, given all pairs of types, the algorithm checks if one is a superset of the other, i.e.,\ they share common fields and methods, and if so, they are related by structural subtyping. 
For e.g., type $empty = []$, will have all classes as structural subtypes. 
%The above algorithm will without differentiating, find all nominal and structural subtypes as we always consider all other classes in the program to be potential subtype candidates.

Note that both approaches to subtyping have their drawbacks.
%However, it can be argued that structural subtyping is counterproductive in writing logically correct programs.
Consider two structurally equivalent record types $coordinate = [x:int, y:int, z:int]$ and $colour = [x:int, y:int, z:int]$. 
While structurally they can be used interchangeably in our type system, their ``meaning'' is completely different.
On the other hand, with nominal subtyping in Java one can define an overridden method to perform the opposite logic to what the super class is expecting. 
Therefore, both approaches leave an expectation on the developer to check what they are doing is correct, hence the integration of both in SFJ is the key to overcoming these drawbacks, as one can choose which approach to use for a given task.

Furthermore, we have extended FJ with method types, demonstrated below to show another possible implementation of the Polygons example \autoref{polygons}.
$$
\syndecl{Diagonal}{Object} {
	\mathbf{int}\ diagonal(\textit{(getDiagonal: void$\rightarrow$int)}  \ p)\{\ldots\}
}
$$

>>>>

Consider the set of polygons, such as triangles, squares and rhombuses given by a class hierarchy.
We want to define a method \emph{diagonal} that takes a polygon and returns the length of its longest diagonal. This method makes sense only if the polygon passed to it has at least four sides, hence triangles are excluded.
In Java this could be implemented in the following ways:
$$
\begin{array}{l}
\synclass Polygon \ \{ \ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Other\_Polygons}{Polygon} {
\\
	\qquad \mathbf{double}\ diagonal(Other\_Polygons \ shape)\ \{\ldots \}
	\\
	\hspace{3.5cm} \ldots
	\\
}
\\
\syndecl{Square}{Other\_Polygons}{\ldots}
\\
\syndecl{Rhombus}{Other\_Polygons} {\ldots}
\end{array}
$$
Or by means of an interface \emph{Diagonal}:
$$
\begin{array}{l}
public \ \syninterface Diagonal\ \{
	\\
	\qquad \mathbf{double}\ diagonal(Polygon \ shape);
	\\
\}
\\
\synclass Polygon \ \{\ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\synidecl{Square}{Polygon} \synimpl Diagonal \ \{\ldots\}
\\
\synidecl{Rhombus}{Polygon} \synimpl Diagonal \ \{\ldots\}
\\
\hspace{4.5cm} \ldots
\end{array}
$$

@@

We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:
$$
\begin{array}{l}
\synclass Polygon \ \{\ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Square}{Polygon} {\ldots}
\\
\syndecl{Rhombus}{Polygon} {\ldots}
\\
\hspace{2.5cm} \ldots
\\
\synclass Diagonal \ \{
	\\
	%\qquad Diagonal ( )\   \{\ super();\ \}
	\hspace{2.5cm} \ldots\\
	\qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
	\\
\}
\end{array}
$$
We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

@@

if all classes are not resolved:
    program contains invalid type definition
\end{verbatim}
%If the types in the program are finite trees whose leaves are constants with no cycles, then at each iteration of the algorithm, either all the types are going to be resolved, or we are going to be able to resolve at least one type. 
%If we do not resolve at least one type and not all types are resolved, we know we have encountered a cycle in the type definition. 
%
Now, we can define the subtyping relation.
Building upon the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the subtyping relation by defining a map from a type to the set of its subtypes, with the property that the set of values of a subtype is included in the set of values of the type.
As a first step, we start with basic types.
\setlength{\abovedisplayskip}{5pt}
\setlength{\belowdisplayskip}{5pt}
\begin{flalign*}
    &Double  = \left\{Double, Float, Int, Short, Byte\right\} && Float = \left\{Float, Short, Byte\right\}\\
    &Long = \left\{Long, Int, Short, Byte\right\}  && Int = \left\{Int, Short, Byte\right\}\\
    &Short = \left\{Short, Byte\right\} \qquad Byte = \left\{Byte\right\} && Boolean = \left\{Boolean \right\} 
\end{flalign*}
%
Note that \emph{Int} is not a subtype of \emph{Float} as a 32-bit \emph{float} cannot represent the whole set of 32-bit \emph{integer} values accurately and therefore \emph{Int} is not fully set-contained, however this is not the case for \emph{Int} and \emph{Double}. 
Similarly, \emph{Long} is not a subtype of \emph{Double}.

@@

function checkSuperSet(class, other):
    boolean flag = true
    for field in class:
        if field contains type not in relation:
            return false
        if other does not have field:
            flag = false
        else:
            if other.field.types not fully contains field.types:
                flag = false
    for method in class:
        if method contains type not in relation:
            return false
        if other does not have method:
            flag = false
        else:
            if other.method.types not fully contains method.types:
                flag = false
    if flag == true:
        add class to other subtype relation
\end{verbatim}
Let us illustrate the subtyping algorithm with our \emph{Polygons} \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.
\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} \qquad Triangle = \left\{Triangle\right\}\\ 
    &Square = \left\{Square\right\} \qquad  Rhombus  = \left\{Rhombus\right\} \qquad
    Diagonal = \left\{Diagonal\right\} 
\end{flalign*}
Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons: 
\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\} 
\end{flalign*}
If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.

The subtyping algorithm finds all nominal and structural subtypes.
It finds all nominal subtypes as they inherit all fields and methods of their super class, so are always guaranteed to be a superset. 
Similarly, given all pairs of types, the algorithm checks if one is a superset of the other, i.e.,\ they share common fields and methods, and if so, they are related by structural subtyping.
For e.g., type $empty = []$, will have all classes as structural subtypes. 
%The above algorithm will without differentiating, find all nominal and structural subtypes as we always consider all other classes in the program to be potential subtype candidates.

Note that both approaches to subtyping have their drawbacks.
%However, it can be argued that structural subtyping is counterproductive in writing logically correct programs.
Consider two structurally equivalent class/record types $coordinate = [x:int, y:int, z:int]$ and $colour = [x:int, y:int, z:int]$. 
While structurally they can be used interchangeably in our type system, their ``meaning'' is completely different.
On the other hand, with nominal subtyping in Java one can define an overridden method to perform the opposite logic to what the super class is expecting. 
Therefore, both approaches leave an expectation on the developer to check what they are doing is correct, hence the integration of both in SFJ is the key to overcoming these drawbacks, as one can choose which approach to use for a given task.

Furthermore, we have extended FJ with method types, demonstrated below to show another possible implementation of the \emph{Polygons} \autoref{polygons}.
$$
\synclass Diagonal \ \{
	\mathbf{int}\ diagonal(\textit{(getDiagonal: void$\rightarrow$int)}  \ shape)\{\ldots\}
\}
$$
